Online Balanced Graph Partitioning: Optimizations for Large Clusters

### 2020-05-07 paper/disc20clustering.tex
Removing Cascade

 \newcommand\mahmoud[1]{\color{orange}\textbf{Mahmoud: #1~}\color{black}}
 \newcommand\stefan[1]{\color{blue}\textbf{\\ Stefan: #1}\color{black}}
 \newcommand\maciek[1]{\color{brown}\textbf{(Maciek: #1)}\color{black}}
 %\newcommand\maciek[1]{}

@@

\section{Bounding the cost of a single reconfiguration of components}

\maciek{Overview of the section, intro. Defer introducing the graph notation, the warm-up does not need it}

\subsection{Warm-up: bounding the reconfiguration for $k=3$}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\textwidth]{figs/substitute}
	\caption{Cluster signatures in warm-up k=3 proof}
\end{figure}

\maciek{Rewrite: cluster types should be cluster signatures}
In our analysis, we distinguish among three types of clusters: $C_1, C_2, C_3$. In a cluster of type $C_i$, the size of the largest component contained in this cluster is $i$.
Before bounding the competitive ratio of \TAlg, we introduce the lemma that estimates the cost of a single repartition of \TAlg.

\begin{lemma}
	\label{lem:1req}
	During a single repartition, \TAlg exchanges at most $2$ pairs of nodes.
\end{lemma}

\begin{proof}
	Observe that when the repartition is triggered by \TAlg, the resulting partitioning is component respecting.
	Otherwise, if it does not exist, \TAlg simply ends the phase and performs no repartition.

	%We distinguish between three types of clusters: $C_1, C_2, C_3$,
	%which we define as follows.
	%A cluster of type $C_1$ the cluster contains $3$ singleton components (this is also the initial configuration of any cluster).
	%A cluster of type $C_2$  contains one component of size $2$ and one component of size $1$.
	%Finally, a cluster of type $C_3$  contains one component of size $3$.

	Consider a request between $u$ and $v$ that triggered the repartition and let $U$ and $V$ be their respective clusters.
	Note that $U\neq V$,
	as otherwise, the request would not trigger a~repartitioning.
	We consider cases based on the types of clusters $U$ and $V$.
	Note that this is impossible that either $U$ or $V$ is of type $C_3$, as otherwise, we merge components of size $3$, and no component respecting partitioning exists, a contradiction.
	If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the reconfiguration is local within $U$ and $V$.
	For two clusters, any configuration can be reached within two swaps, due to the fact that clusters are indistinguishable.

	Finally, we consider the case where both $U$ and $V$ are of type $C_2$. Note that $(u,v)$ cannot both belong to a component of size $2$, as this would mean that \TAlg has the component of size $4$, a contradiction with the case assumption that the component respecting partitioning exists. 
	Otherwise, if either $u$ or $v$ belongs to a component of size $2$, then it suffices to exchange components of size $1$ between $U$ and $V$.
	Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a cluster different from $U$ and $V$.
	Note that in such case, a $C_1$-type cluster $W$ exists, as otherwise no component respecting partitioning exists. In this case \TAlg performs one swap, i.e., it exchanges the nodes $u$ and $v$ with any two nodes of $W$.
\end{proof}

Corollary: DET2 is $O(\ell)$-competitive for $k=3$.

\subsection{Characterization of the migration graph}

Consider a component merge action performed by \TAlg that triggers a reconfiguration from a configuration $C_I$ to a configuration $C_F$.
To execute the reconfiguration, the nodes of two non-collocated components migrate to a common cluster.
Each cluster contains exactly $k$ nodes, thus other nodes migrate to take the place of nodes of merged components.
Their place must be occupied, too, and this way a single merge action may trigger migrations in clusters not directly involved in a merge action (for an example, see Section~\ref{ssec:example}). 
Recall that both $C_I$ and $C_F$ are component-respecting configurations, hence a migration of a single node entails migrations of all nodes of its cluster.

In Section~\ref{ssec:cascade}, we upper bound the cost incurred by a reconfiguration that follows a single component merge.
To this end, we introduce a \emph{migration graph} that models the reconfiguration.
In the following, we characterize the structure of the migration graph.

\noindent
\textbf{Vertices of the migration graph and the core vertices.}
The vertices of the migration graph are all $\ell$ clusters of the instance.
We distinguish the \emph{core vertices} that correspond to clusters directly involved in the merge operation: the clusters containing the to-be-merged components in $C_I$ and the cluster containing the merged component in $C_F$.
There are at most $3$ core vertices in each migration graph (at most two components participate in the merge, and these may migrate to a third cluster).

\noindent
\textbf{Edges and their labels.}
The edges of the migration graph denote the migration of nodes from $C_I$ to $C_F$.
Each edge is labeled with a number $m$ that denotes the number of migrated nodes between clusters.
Each edge is directed from the cluster that contained the nodes to the cluster they migrated to.
Both edges might exists between a pair of nodes, as whole components of nodes migrate, and the back-and-forth exchange may be needed for the configuration $C_F$ to be component-respecting.

\noindent
\textbf{Vertex degree.}
At most $k$ nodes may migrate from a cluster and at most $k$ nodes may migrate into a cluster.
Thus, the sum of the labels for both ingoing and outgoing edges of each vertex is at most $k$.
This implies that the number of ingoing and outgoing edges is also bounded by $k$.

\noindent
\textbf{Flow preservation.}
In any feasible configuration, each cluster contains exactly $k$ nodes.
Thus, for any vertex, the sum of labels of ingoing edges must equal the sum of labels of outgoing edges.
\maciek{Can we say that this implies that the graph is a circulation?}

\noindent
\textbf{Migration graph and the cost of reconfiguration.}
The cost of cluster reconfiguration equals the number of exchanged nodes multiplied by $\alpha$.
From the standpoint of a migration graph, this corresponds to the sum of labels on edges of the graph.
A trivial upper bound on the cost is the size of the connected component in the migration graph multiplied by $2\alpha k$.

Note that for a single component merge, multiple migration graphs may exist, and multiple of them might have the optimal cost.
In Section \ref{ssec:cascade}, we show that among all migration graphs for given component merge, there exists an optimal graph with the bounded cycle length.
Combined with other properties, this allows to bound the size of the migration graph, and consequently bound the cost of each repartition, that results in a bound of the competitive ratio for \TAlg.

\subsection{Costly cluster reconfiguration example}
\label{ssec:example}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\textwidth]{figs/substitute}
	\caption{A costly cascade example. On the left: the configuration prior to the merge and the merge of two components (denoted solid). On the right: corresponding migration graph for the optimal (unique) reconfiguration.}
\end{figure}

In this section we present an example of a merge action that results in a $k/3$(?) length cycle in an optimal migration graph.

\subsection{Bounds on the size of a migration graph}
\label{ssec:cascade}

In \cite{repartition-disc}, authors trivially bounded the reconfiguration cost for each merge action by $k \cdot \ell$.
This roughly corresponds to migrating every node in every cluster.
Now, we bound the cost of a single repartition of \TAlg as a function of $k$.
Note that the bound of $k \cdot \ell$ is still valid, and the resulting bound is the minimum of these two.

We begin, by bounding the length of each cycle in one of optimal solutions.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\textwidth]{figs/substitute}
	\caption{Illustration to bounding the cycle length.}
	\label{fig:cascade-illustration}
\end{figure}

\begin{theorem}
	\label{th:cascade-cycles}
	Consider a single merge action of \TAlg.
	There exists an optimal cost reconfiguration with the migration graph, where each cycle have length at most $k$.
\end{theorem}

\begin{proof}
	Assume the contrary, i.e., for each optimal cluster reconfiguration, its migration graph contains a cycle at least $k+1$.

	Fix a reconfiguration and its migration graph $G$ that has the shortest length $L$ of the longest cycle, and has the least number of cycles of length $L$.
	Let $p = \lbrace v_1, v_2, \ldots, v_L \rbrace$ be any cycle of length $L$.
	As edge labels are from $\{ 1, 2, \ldots, k \}$, there exists a pair of edges $e, f$ with equal labels.

	We show that in such case, we can modify the solution without affecting its cost and feasibility to decrease the number of cycles of length $L$ by $1$.
	This would contradict the assumption that we consider a solution with the minimal number of cycles of length $L$.

	We construct a new migration graph $G'$ of equal cost by swapping the (destination) endpoints of $e$ and $f$ (cf. Figure~\ref{fig:cascade-illustration})
	Formally, let $\lbrace a, b \rbrace = e$ and $\lbrace c, d \rbrace = f$.
	We remove $e$ and $f$ and add $e' = \lbrace a, d \rbrace$ and $f' = \lbrace c, d \rbrace$.
	This way we produced a feasible solution.

	It remains to show that while decreasing the number of cycles of length $L$, we did not created any new cycles of length at least $L$.
	Assume that adding $e'$ created a new cycle of length at least $L$.
	However, $a$ and $d$ were already connected in $G$ by a directed path of length at least $2$, and a cycle strictly longer than $L$ existed in $G$.

	Assume that adding $f'$ created a new cycle $s^*$ of length $L' \geq L$.
	Contrary to the previous case, the directed cycle from $e$ to $b$ might not exist in $G$.
	However, this would mean that $b$ and $e$ were connected by a path of length $L'-1$.
	This means that we might have followed this alternative path from $b$ to $e$ to obtain a cycle strictly longer than $L$ (cf. Figure~\ref{fig:cascade-illustration}).

\begin{theorem}
	Consider a single repartition of DET2.
	The number of clusters it involves is bounded by $k^k$.
	\label{th:cascade}
\end{theorem}

\begin{proof}
	Each node has degree at most $k$.
	Every subset has flow conservation property, hence each cycle traverses one of the nodes of graph origin/core. (no subsets with just incoming edges to the origin part)
	Combine with Theorem \ref{th:cascade-cycles}.

Corollary: Algorithm DET2 is $O(\ell \cdot k \cdot \min \{ \ell, k^k \})$-competitive.

>>>>

 \newcommand\mahmoud[1]{\color{orange}\textbf{Mahmoud: #1~}\color{black}}
 \newcommand\stefan[1]{\color{blue}\textbf{Stefan: #1}\color{black}}
 \newcommand\maciek[1]{\color{brown}\textbf{(Maciek: #1)}\color{black}}
 %\newcommand\mahmoud[1]{}
 %\newcommand\stefan[1]{}
 %\newcommand\maciek[1]{}

@@
